---
title: "Multiple Testing When Many p-Values are Uniformly Conservative"
subtitle: "Multiple Comparisons and Selective Inference Sem. A 2020"
author: "Giora Simchoni"
institute: "Stat. and OR Department, TAU"
date: "`r Sys.Date()`"
output_dir: "images"
output:
  xaringan::moon_reader:
    css: "slides.css"
    seal: false
    includes:
      in_header: "header.html"
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

class: title-slide

## Multiple Testing When Many P-Values are Uniformly Conservative

#### Qingyuan Zhao, Dylan S. Small, and Weijie Su (2019)

##### Presented by: Giora Simchoni

### Multiple Comparisons and Selective Inference Sem. A 2020

#### Stat. and OR Department, TAU
#### 2020-11-05

---
```{r child = "setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

class: section-slide

# At Highest Level

---

## There's a Story

.no_shadow[
<center><img src = "images/high_level_story.png" style="width: 100%"></center>
]

---

class: section-slide

# Motivation

---

## Educational Intervention

[Cooper et al. (2003)](https://www.jstor.org/stable/3516042) Meta-Analysis:

<center><img src = "images/cooper_abstract.png" style="width: 67%"></center>

---

### A typical (over-simplified) Mixed Model

$y_i = \beta_0 + \beta_MX_i + b_S + \epsilon_i$

Where:

$y_i$: Student $i$ "achievement"

$$X_i = \begin{cases}
      0, & \text{Student i in Modified Calendar}\ \\
      1, & \text{otherwise}
    \end{cases}$$

$b_S \sim N(0, \sigma_S^2)$ (Random Effect)

And we're much interested in:

$H: \beta_M \leq 0 \space vs. \space K: \beta_M > 0$

---

## Result: "quite small"

<center><img src = "images/cooper_results.png" style="width: 43%"></center>
